---
title: "Statistical Analysis of Short-Read vs. Long-Read Correlation Factors"
author: "Antigravity Pair Programmer"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(lme4)
library(lmerTest)

# Helper function to compute pseudo-R2 (Nakagawa's R2) when MuMIn is not available
get_r2 <- function(model) {
  if (requireNamespace("MuMIn", quietly = TRUE)) {
    return(MuMIn::r.squaredGLMM(model))
  } else {
    # Manual Nakagawa's pseudo-R2 calculation for random intercept models
    fix_var <- var(predict(model, re.form = NA))
    vc <- as.data.frame(VarCorr(model))
    rand_var <- sum(vc$vcov[vc$grp != "Residual"])
    resid_var <- vc$vcov[vc$grp == "Residual"]
    total_var <- fix_var + rand_var + resid_var
    r2_marginal <- fix_var / total_var
    r2_conditional <- (fix_var + rand_var) / total_var
    
    res <- matrix(c(r2_marginal, r2_conditional), nrow = 1)
    colnames(res) <- c("R2m", "R2c")
    rownames(res) <- "theoretical"
    return(res)
  }
}
```

# Introduction

We analyze the factors that impact the Spearman correlation between short-read (Illumina) and long-read expression estimates at the transcript level. Following the approach developed for the Mouse dataset, we model the Fisher-z transformed correlation coefficients as a function of processing tool (`Tool`), normalization method (`Normalization`), and sequencing technology (`Technology`), with individual samples modeled as a random effect to account for replicate and sample-specific correlations:

$$ \text{Correlation}_z \sim \text{Tool} \times \text{Normalization} + \text{Tool} \times \text{Technology} + \text{Normalization} \times \text{Technology} + (1 \mid \text{Sample}) $$

---

# 1. Mouse (NIH) Dataset Analysis

```{r mouse_analysis}
# Load mouse correlation dataset
mouse_dat <- readRDS("/home/alejandro/Desktop/bias_paper/LMM/sr_vs_lr/mouse_sr_correlations.rds")

# Prepare factors
mouse_dat$Technology <- as.factor(mouse_dat$Dataset)
mouse_dat$Technology <- relevel(mouse_dat$Technology, ref = "ont")
mouse_dat$Tool <- as.factor(mouse_dat$Tool)
mouse_dat$Tool <- relevel(mouse_dat$Tool, ref = "bambu")
mouse_dat$Normalization <- as.factor(mouse_dat$Normalization)
mouse_dat$Normalization <- relevel(mouse_dat$Normalization, ref = "CPM")
mouse_dat$Correlation_z <- atanh(mouse_dat$Correlation)

# Fit mixed-effects model
model_mouse <- lmer(Correlation_z ~ Tool * Normalization + Tool * Technology + Normalization * Technology + 
                     (1 | Sample), data = mouse_dat)

# ANOVA (Type III)
anova(model_mouse, type = 3)

# Model summary (coefficients)
summary(model_mouse)

# Variance explained (R2)
get_r2(model_mouse)
```

---

# 2. Human Induced Neurons (Neuron) Dataset Analysis

```{r neuron_analysis}
# Load neuron correlation dataset
neuron_dat <- readRDS("/home/alejandro/Desktop/bias_paper/LMM/sr_vs_lr/neuron_sr_correlations.rds")

# Prepare factors
neuron_dat$Technology <- as.factor(neuron_dat$Dataset)
neuron_dat$Technology <- relevel(neuron_dat$Technology, ref = "ont")
neuron_dat$Tool <- as.factor(neuron_dat$Tool)
neuron_dat$Tool <- relevel(neuron_dat$Tool, ref = "bambu")
neuron_dat$Normalization <- as.factor(neuron_dat$Normalization)
neuron_dat$Normalization <- relevel(neuron_dat$Normalization, ref = "CPM")
neuron_dat$Correlation_z <- atanh(neuron_dat$Correlation)

# Fit mixed-effects model
model_neuron <- lmer(Correlation_z ~ Tool * Normalization + Tool * Technology + Normalization * Technology + 
                      (1 | Sample), data = neuron_dat)

# ANOVA (Type III)
anova(model_neuron, type = 3)

# Model summary (coefficients)
summary(model_neuron)

# Variance explained (R2)
get_r2(model_neuron)
```

---

# 3. LongBench Dataset Analysis

```{r longbench_analysis}
# Load longbench correlation dataset
longbench_dat <- readRDS("/home/alejandro/Desktop/bias_paper/LMM/sr_vs_lr/longbench_sr_correlations.rds")

# Prepare factors
longbench_dat$Technology <- as.factor(longbench_dat$Dataset)
longbench_dat$Technology <- relevel(longbench_dat$Technology, ref = "ont")
longbench_dat$Tool <- as.factor(longbench_dat$Tool)
longbench_dat$Tool <- relevel(longbench_dat$Tool, ref = "bambu")
longbench_dat$Normalization <- as.factor(longbench_dat$Normalization)
longbench_dat$Normalization <- relevel(longbench_dat$Normalization, ref = "CPM")
longbench_dat$Correlation_z <- atanh(longbench_dat$Correlation)

# Fit mixed-effects model
model_longbench <- lmer(Correlation_z ~ Tool * Normalization + Tool * Technology + Normalization * Technology + 
                         (1 | Sample), data = longbench_dat)

# ANOVA (Type III)
anova(model_longbench, type = 3)

# Model summary (coefficients)
summary(model_longbench)

# Variance explained (R2)
get_r2(model_longbench)
```

---

# 4. Combined Coefficient Heatmap

We summarize the regression coefficients and statistical significance from all three models. 
The colors represent the p-value classification, and the values within the cells show the regression coefficient (Estimate).

```{r combined_heatmap, fig.width=7, fig.height=10}
# Clean term names helper
clean_term <- function(term) {
  term <- gsub("^Toolflair$", "flair (vs. Bambu)", term)
  term <- gsub("^Toolisoquant$", "isoquant (vs. Bambu)", term)
  term <- gsub("^Toolkallisto$", "kallisto (vs. Bambu)", term)
  term <- gsub("^Tooloarfish$", "oarfish (vs. Bambu)", term)
  
  term <- gsub("^Normalizationcqn$", "cqn (vs. CPM)", term)
  term <- gsub("^NormalizationEDA$", "EDA (vs. CPM)", term)
  term <- gsub("^Normalizationratio_counts$", "SMR (vs. CPM)", term)
  term <- gsub("^NormalizationTPM$", "TPM (vs. CPM)", term)
  
  term <- gsub("^Technologyisoseq$", "IsoSeq (vs. cDNA ONT)", term)
  term <- gsub("^Technologykinnex$", "Kinnex (vs. cDNA ONT)", term)
  term <- gsub("^Technologyont_drna$", "dRNA ONT (vs. cDNA ONT)", term)
  
  if (grepl(":", term)) {
    parts <- strsplit(term, ":")[[1]]
    clean_parts <- sapply(parts, function(p) {
      p <- gsub("^Tool", "", p)
      p <- gsub("^Normalization", "", p)
      p <- gsub("^Technology", "", p)
      if (p == "ratio_counts") p <- "SMR"
      if (p == "ont_drna") p <- "dRNA ONT"
      if (p == "isoseq") p <- "IsoSeq"
      if (p == "kinnex") p <- "Kinnex"
      p
    })
    return(paste(clean_parts, collapse = " \u00D7 "))
  }
  return(term)
}

# Assign term types helper
get_term_type <- function(term) {
  if (!grepl(":", term)) {
    if (grepl("^Tool", term)) return("Tool Effects")
    if (grepl("^Normalization", term)) return("Normalization Effects")
    if (grepl("^Technology", term)) return("Technology Effects")
    return("Main Effects")
  } else {
    parts <- strsplit(term, ":")[[1]]
    has_tool <- any(grepl("^Tool", parts))
    has_norm <- any(grepl("^Normalization", parts))
    has_tech <- any(grepl("^Technology", parts))
    
    if (has_tool && has_norm) return("Tool \u00D7 Normalization")
    if (has_tool && has_tech) return("Tool \u00D7 Technology")
    if (has_norm && has_tech) return("Normalization \u00D7 Technology")
    return("Interactions")
  }
}

# Extract coefs function
get_model_coefs <- function(model, dataset_name) {
  coefs <- summary(model)$coefficients
  df <- data.frame(
    Term = rownames(coefs),
    Estimate = coefs[, "Estimate"],
    PValue = coefs[, "Pr(>|t|)"],
    DatasetName = dataset_name,
    stringsAsFactors = FALSE
  )
  return(df)
}

# Extract coefficients for all three models
coefs_mouse <- get_model_coefs(model_mouse, "Mouse")
coefs_neuron <- get_model_coefs(model_neuron, "iNeuron")
coefs_longbench <- get_model_coefs(model_longbench, "LongBench")

# Combine and process
all_coefs <- rbind(coefs_mouse, coefs_neuron, coefs_longbench) %>%
  filter(Term != "(Intercept)")

all_coefs$CleanTerm <- sapply(all_coefs$Term, clean_term)
all_coefs$TermType <- sapply(all_coefs$Term, get_term_type)

term_type_order <- c(
  "Tool Effects",
  "Normalization Effects",
  "Technology Effects",
  "Tool \u00D7 Normalization",
  "Tool \u00D7 Technology",
  "Normalization \u00D7 Technology"
)
all_coefs$TermType <- factor(all_coefs$TermType, levels = term_type_order)
all_coefs$DatasetName <- factor(all_coefs$DatasetName, levels = c("Mouse", "LongBench", "iNeuron"))

all_coefs <- all_coefs %>%
  mutate(
    PValCat = case_when(
      PValue < 0.05 & Estimate > 0 ~ "Significant (Positive)",
      PValue < 0.05 & Estimate < 0 ~ "Significant (Negative)",
      TRUE                         ~ "NS (p \u2265 0.05)"
    ),
    PValCat = factor(PValCat, levels = c("Significant (Positive)", "NS (p \u2265 0.05)", "Significant (Negative)"))
  )

# Expand combinations to ensure complete grid with grey placeholder for missing levels
all_combinations <- expand.grid(
  CleanTerm = unique(all_coefs$CleanTerm),
  DatasetName = c("Mouse", "LongBench", "iNeuron"),
  stringsAsFactors = FALSE
)

term_details <- all_coefs %>% select(CleanTerm, TermType) %>% distinct()
all_combinations <- all_combinations %>% left_join(term_details, by = "CleanTerm")

heatmap_data <- all_combinations %>%
  left_join(all_coefs %>% select(CleanTerm, DatasetName, Estimate, PValue, PValCat), by = c("CleanTerm", "DatasetName"))

heatmap_data$DatasetName <- factor(heatmap_data$DatasetName, levels = c("Mouse", "LongBench", "iNeuron"))
heatmap_data$PValCat[is.na(heatmap_data$PValCat)] <- "NS (p \u2265 0.05)"

heatmap_data <- heatmap_data %>%
  mutate(
    EstimateText = ifelse(is.na(Estimate), "", sprintf("%.3f", Estimate))
  )

# Diverging color palette mapping (Red for Positive, Blue for Negative significance)
colors_signif <- c(
  "Significant (Positive)" = "#DE2D26", 
  "NS (p \u2265 0.05)"       = "#F0F0F0",
  "Significant (Negative)" = "#3182BD"
)

# theme_nature from fig1.Rmd
theme_nature <- function(base_size = 6.5, base_family = "sans") {
  theme_classic(base_size = base_size, base_family = base_family) +
    theme(
      axis.line = element_line(linewidth = 0.35, colour = "black"),
      axis.ticks = element_line(linewidth = 0.35, colour = "black"),
      axis.title = element_text(size = base_size),
      axis.text = element_text(size = base_size - 0.5, color = "black"),
      legend.title = element_text(size = base_size - 0.3),
      legend.text = element_text(size = base_size - 0.7),
      strip.text = element_text(size = base_size - 0.3, face = "bold"),
      plot.title = element_text(size = base_size + 0.5, face = "bold"),
      panel.grid = element_blank()
    )
}

p_heatmap <- ggplot(heatmap_data, aes(x = DatasetName, y = CleanTerm, fill = PValCat)) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(aes(label = EstimateText), size = 1.8, color = "black") +
  facet_grid(TermType ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(
    values = colors_signif,
    name = "Significance",
    na.value = "#F0F0F0"
  ) +
  labs(
    x = "Dataset",
    y = "Regression Coefficient (Effect on Correlation)"
  ) +
  theme_nature() +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_text(size = 7, face = "bold"),
    strip.text.y = element_text(angle = 0, face = "bold", hjust = 0, size = 6.5),
    legend.position = "right",
    legend.title = element_text(size = 6.5, face = "bold"),
    legend.text = element_text(size = 5.8),
    legend.key.size = unit(3, "mm")
  )

# Display in Rmd
p_heatmap

# Export publication-quality figures
save_pub_r <- function(plot, filename, width_mm = 180, height_mm = 240, dpi = 600) {
  w <- width_mm / 25.4
  h <- height_mm / 25.4
  
  svglite::svglite(paste0(filename, ".svg"), width = w, height = h)
  print(plot)
  dev.off()
  
  grDevices::cairo_pdf(paste0(filename, ".pdf"), width = w, height = h, family = "sans")
  print(plot)
  dev.off()
  
  ragg::agg_tiff(paste0(filename, ".tiff"), width = w, height = h, units = "in", res = dpi)
  print(plot)
  dev.off()
  
  ggsave(paste0(filename, ".png"), plot, width = w, height = h, dpi = 150)
}

save_pub_r(p_heatmap, "regression_coefficients_heatmap", height_mm = 240)
```

